Metric Spaces with Expensive Distances

### 2019-04-24 paper/expensive-distances-main.tex
some Michaels' changes

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of all pairwise distances or runs slighly slower, depending on the approximation
quality.

@@

We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centres uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centres. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In addition to the standard Euclidean norm on $\RR^n$, we used also $\ell_1$ and $\ell_\infty$ norms.
In all these experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

@@

The realistic data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with \libraryname{DIPHA}
library \cite{bauer2014distributed}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
\grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
any advantage of our algorithm disappears. However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
We can offer the following explanation\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}}. For higher values of $q$,
\label{}
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. \ignore{On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; if we are happy with a 1.5 approximation, then we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in the brute-force columns
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, or
it can be computed with the user-given accuracy with the auction algorithm,
as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
To make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, by a factor of 2 for $\wsdist{1}$.

>>>>

We propose several heuristics of how to explore the metric space to obtain accurate lower and upper bounds
with a small number of distance computation. Once the ratio of upper and lower bound is at most $(1+\eps)$
for each point pair, the set of all computed distances forms the spanner.
The experimentally most successful exploration strategy that we found is to
repeatedly query the distance of a pair with the worst ratio of upper and lower bound.
We call the obtained spanner the \emph{blind greedy spanner}, as opposed to the well-known
\emph{greedy spanner} that precomputes all pairwise distances and only maintains upper bounds (\cite{althofer1993sparse}).
Remarkably, we were not able to improve the quality when knowing initial $2$-approximations of all point pairs.
We also compare with a spanner construction based on WSPD. Our simple algorithms tend to give much smaller
spanners on the tested examples. Nevertheless, we leave the question open whether our construction
yields a spanner of asymptotically linear size. From the practical point of view, our algorithm either
outperforms naive computation of all pairwise distances or runs slightly slower, depending on the approximation
quality.

@@

We tested the algorithms for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centres uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centres. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{\xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In addition to the standard Euclidean norm on $\RR^n$, we used also $\ell_1$ and $\ell_\infty$ norms.
In all these experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

@@

The realistic data sets that we used
are the persistence diagrams
computed from the McGill shape benchmark \cite{zhang2005mcgill}.
More precisely, we generated two sets of diagrams,
\dtype{original} and \dtype{perturbed}. The \dtype{original}
diagrams were computed on the points of the 450 original
point clouds from the shape dataset; we only used
dimension 0, since the diagrams in this dimension are larger in size.
We also added random noise to points of the original shapes,
and computed persistence diagrams in dimension 0 of these modified
shapes; these diagrams form the \dtype{perturbed} dataset.
Persistence diagrams were computed with the \libraryname{DIPHA}
library \cite{bauer2014distributed}.

Sparseness values of \bgrdy spanner on the whole \dtype{original} dataset
for different values of $\eps$ and $q$ are presented in \cref{tbl:mcgill_original_blind_greedy_spanner_sparseness};
the \grdy algorithm is used as baseline.  The first thing to note is that for higher values of $q$
\footnote{definition of $\wsdist{q}$ can be found in \cref{sec:appendix_background}} 
any advantage of our algorithm disappears. A plausible explanation 
is that for higher values of $q$, the doubling dimension becomes larger;
this is confirmed by the increase in the size of the greedy spanner.
\ignore{
However, the cardinality of the \grdy
spanner is also increasing; for instance, if we want to achieve 10 \% accuracy,
then even the \grdy spanner is almost half of the complete graph.
longer edges of an optimal matching play a more important role
than shorter ones; as $q$ tends to $\infty$, Wasserstein distance
approaches bottleneck distance $\btdist$, which depends on the longest
edge only. If all diagrams come from the same dataset,
then there is a large number of pairs of diagrams whose optimal matchings
have the longest edges of approximately the same length. This fact means that we are in the case
of high doubling dimension, being close to a clique, and there is no hope
of getting a small spanner. On the other hand, computation of $\btdist$ in practice
requires significantly less time than computation of $\wsdist{q}$, and the case
of $\wsdist{1}$ and $\wsdist{2}$ is of practical importance.}
For $\wsdist{1}$ we observe more encouraging figures, like avoiding 70 \% of distance
computations for $\eps = 0.1$; for a 1.5 approximation, we
reduce the number of distances computed by a factor of 10.

We also compare the time that is needed to compute the matrix of all pairwise distances
naively with the time that is needed to approximate it with \bgrdy spanner.
The timings are given in \cref{tbl:mcgill_original_blind_greedy_spanner_timing}.
First of all, we should explain why the timings in the brute-force columns
depend on $\eps$. Wasserstein distance can either be computed exactly,
by using the Hungarian algorithm, or
it can be computed with the user-given accuracy with the auction algorithm,
as in \hera. Exact computation is very expensive \cite{kerber2017geometry}, and in our code for experiments
with spanner algorithms we always compute $\wsdist{q}$ with relative error 0.01.
To make the comparison absolutely fair, when we compute a \espanner,
we compare its timing with the timing of \hera called for each pair of diagrams with relative error $\eps$.
Let us emphasize that the diagrams we have are rather small, with average cardinality
350 points. Nevertheless, on this data set \bgrdy spanner
provides a substantial improvement in terms of real running time, and we expect this advantage
to grow for diagrams of larger cardinality.
